---
output: github_document
editor_options: 
  markdown: 
    wrap: 80
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# review

<!-- badges: start -->

[![lifecycle](https://lifecycle.r-lib.org/articles/figures/lifecycle-experimental.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Project Status:
WIP](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![devel-version](https://img.shields.io/badge/devel%20version-0.0.2-blue.svg)](https://github.com/dataobservatory-eu/review/tree/devel)
[![dataobservatory](https://img.shields.io/badge/ecosystem-dataobservatory.eu-3EA135.svg)](https://dataobservatory.eu/)
[![R-CMD-check](https://github.com/dataobservatory-eu/review/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/dataobservatory-eu/review/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/dataobservatory-eu/review/graph/badge.svg)](https://app.codecov.io/gh/dataobservatory-eu/review)
<!-- badges: end -->

**review** a tidy relational algebra and a dual extension of the *tidyverse* data model for iterative semantic review of tabular data.

The package provides a vectorised review workspace for creating, reviewing, documenting, and finalising semantic claims while remaining compatible with ordinary data frames. Rather than prescribing how reviews are carried out, it supports reproducible review workflows that integrate naturally with external tools such as spreadsheets, databases, and interactive review interfaces.

Although it is not an ontology toolkit, its design is informed by provenance, policy, and statistical quality-control models, allowing review histories to be audited, serialised, compared with expected review activities, and exchanged in a reproducible form.

## Installation

Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("dataobservatory-eu/review")
```

## Review workflow

The review package provides four verbs for reproducible semantic review.

``` text
claims_df()
      ↓
review()
      ↓
explain()
      ↓
finalise_review()
```

- `claims_df()` creates reviewable claims from an ordinary data frame.

<!-- -->

- `review()` allocates a review round and labels the intended review task.

<!-- -->

- `explain()` records how the review was carried out, including provenance and
  optional reviewer comments.

<!-- -->

- `finalise_review()` accepts the reviewed values as the current candidate
  claims while preserving the complete review history.

## Example

```{r example}
library(review)

claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
)

reviewed <- claims |>
  review(
    "circumference",
    label = "Remeasure the circumference of each tree."
  )

reviewed$circumference_review_1[1] <- 31

reviewed <- reviewed |>
  explain(
    activity = "manual review",
    agent = person("Jane", "Doe", role = "rev"),
    used = "doi:10.5281/zenodo.1234567",
    comment = "Verified against the laboratory notebook."
  ) |>
  finalise_review()

reviewed
```

## Learn more

The package includes two introductory vignettes:

- [The Review Algebra](https://review.dataobservatory.eu/articles/intro.html)
  introduces the four review verbs and demonstrates a complete review workflow.

- [From Review Algebra to Provenance   Modelling](https://review.dataobservatory.eu/articles/prov.html) explains the underlying provenance model and its relationship to the W3C PROV data model.

You can read them by clicking to the links to the package website, or, after installing the package, open them with:

```{r, eval=FALSE}
vignette("review")
vignette("prov")
```


## Contributing

Contributions are welcome.

Please read the following documents before submitting pull requests:

- **Contributor Covenant Code of Conduct**

- **Contributing Guidelines** (`CONTRIBUTING.md`)

- **Coding Guidelines** (`CODING-GUIDELINES.md`)

## Project links

- Website: <https://review.dataobservatory.eu/>

- Issues: <https://github.com/dataobservatory-eu/review/issues>
