---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# review

<!-- badges: start -->

[![lifecycle](https://lifecycle.r-lib.org/articles/figures/lifecycle-experimental.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Project Status:
WIP](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![devel-version](https://img.shields.io/badge/devel%20version-0.0.0090-blue.svg)](https://github.com/dataobservatory-eu/fscontext/tree/devel)
[![dataobservatory](https://img.shields.io/badge/ecosystem-dataobservatory.eu-3EA135.svg)](https://dataobservatory.eu/)

<!-- badges: end -->

**review** provides a tidy relational algebra for iterative semantic review of
tabular data.

Rather than overwriting reviewed values, the package creates successive review
rounds that preserve candidate values, review history, reviewer provenance, and
observation status. It is designed for reproducible review workflows in official
statistics, digital humanities, archives, and other data-intensive disciplines.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("dataobservatory-eu/review")
```

## Review workflow

The review algebra consists of four operations.

```text
claims_df()
      ↓
review()
      ↓
explain()
      ↓
finalise_review()
```

- `claims_df()` creates reviewable claims from an ordinary data frame.
- `review()` allocates one or more review rounds.
- `explain()` records provenance for each review round.
- `finalise_review()` promotes the current review into the next candidate
  version.

## Example

```{r}
library(review)

claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
)

reviewed <- claims |>
  review("circumference")

reviewed$circumference_review_1[1] <- 31

reviewed <- reviewed |>
  explain(
    activity = "manual review",
    agent = person("Jane", "Doe", role = "rev"),
    used = "doi:10.5281/zenodo.1234567"
  ) |>
  finalise_review()

reviewed
```

For a complete introduction, see the **Review Algebra** vignette.

- Website: <https://review.dataobservatory.eu/>
- Issues: <https://github.com/dataobservatory-eu/review/issues>
