---
title: "The Review Algebra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Review Algebra}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(review)
```

## Motivation

Review is a ubiquitous activity in official statistics, digital humanities, archives, machine learning, and data engineering. Existing workflows often rely on spreadsheets or interactive tools that make review difficult to reproduce.

## Claims

Introduce

`claims_df()`

Explain

scope
subject
reviewable variables
## Review rounds


Introduce

`review()`

Explain that each review round creates a new semantic version of one or more claims.

Illustrate

```
candidate
↓
review_1
↓
review_2
```

## Explanations

Introduce `explain()`

Explain that it records provenance for the review round rather than modifying data.

## Finalisation

Show

finalise_review()

mapping the reviewed values back into a release.

## Example

The Orange example followed by the Garamantas example.
