---
title: "The Review Algebra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Review Algebra}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(review)
```

## Motivation

Many data workflows require human review before publication. Official
statistics, digital humanities, archives, machine learning, and data engineering
all rely on iterative assessment of candidate values before they become part of
a released dataset.

The goal of **review** is to make semantic review explicit and reproducible.
Rather than overwriting existing values, each review round creates a new version
of one or more reviewable claims while preserving the previous versions and
their provenance. This review history can be inspected, reproduced, and extended
with additional review rounds.

## Creating reviewable claims

A review begins with a collection of claims.

```{r claims_df}
claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
)

head(claims, n = 6)
```

A `claims_df` separates variables into three structural roles and one or more
reviewable variables.

- `Identifier` uniquely identifies each claim.
- `Scope` defines the context in which the claim is made.
- `Subject` identifies the entity being described.
- `Reviewable` variables contain the values that may change during semantic
  review.

The structural variables remain stable throughout the review process, while
reviewable variables may acquire successive reviewed versions.

Readers familiar with the tidy data principles will recognise this separation.
Structural variables play a role analogous to identifiers, dimensions, and
attributes in statistical data production, while the reviewable variables
correspond to measured values that are iteratively reviewed and improved. The
structural variables provide the context for interpretation and allow claims to
be grouped, filtered, or compared, whereas the reviewable variables capture the
semantic content that evolves during the review process.

```{r names}
names(claims)
```

Notice that `circumference` has become `circumference_candidate`. Candidate
values are the current working version of each reviewable variable and form the
starting point for the first review round.

## First review round

A review round allocates one or more review columns. At this stage the package
records that a review is about to take place, but it does not prescribe how that
review should be performed.

```{r review1}
reviewed <- claims |>
  review("circumference")

names(reviewed)
```

The review algebra is independent of the review interface. Reviewers may use
whatever environment best suits the task, provided that the reviewed values are
written back into the allocated review columns.

The package deliberately does not prescribe how the review is carried out. For
example, the review may be

- a `dplyr::mutate()` pipeline,
- values imported from a reviewed CSV file,
- an interactive R console session,
- a Shiny application,
- an Excel or LibreOffice spreadsheet,
- an OpenRefine workflow.

For illustration we edit one value directly.

```{r review1b}
reviewed$circumference_review_1[1] <- 31
```

Once the review values have been entered, the review round can be documented.

```{r explain1}
reviewed <- reviewed |>
  explain(
    activity = "manual review",
    agent = person(
      "Jane",
      "Doe",
      role = "rev"
    ),
    used = "doi:10.5281/zenodo.1234567"
  )
```

`explain()` records how the review was performed, who performed it, and\
what evidence or resources were used. Provenance is stored separately from\
the reviewed values, allowing the same review data to be interpreted or\
exported in different provenance models.

Readers familiar with FAIR data principles, reproducible research, or\
statistical and digital heritage workflows may recognise a common problem:\
as data are cleaned, harmonised, and reviewed, an increasing share of the\
knowledge about *why* particular decisions were made remains only in the\
analyst's or curator's head. The resulting datasets may be reusable, but\
the review process itself is often difficult to inspect, reproduce, or\
audit.

The review algebra aims to record as much of this review history as\
possible while requiring as little additional documentation effort as\
possible. By separating reviewed values from the provenance of the review,\
it records which **agents** (people, software, or AI models) performed\
which **activities**, and what **entities** (datasets, publications,\
files, or other resources) informed those decisions.

This lightweight provenance layer follows the concepts of the PROV data\
model without requiring users to work directly with RDF or ontologies. It\
can later be serialised to PROV-O or aligned with standards such as SDMX,\
DataCite, or archival provenance models. The objective is to improve the\
**reviewability**, **auditability**, **reproducibility**, and ultimately\
the **trustworthiness** and **reusability** of reviewed data while\
remaining compatible with ordinary R data frames.

```{r attributes1}
attr(reviewed, "prov_activity")

attr(reviewed, "prov_agent")

attr(reviewed, "prov_used")
```

## Second review round

Additional review rounds continue from the current accepted review rather than
from the original candidate values. Each round therefore forms a chain of
semantic revisions.

```{r review2}
reviewed <- reviewed |>
  review("circumference")
```

Again, the review itself may take place using any suitable workflow.

```{r review2value}
reviewed$circumference_review_2[1] <- 32
```

Each review round has its own provenance. Different reviewers, software, or
evidence can therefore contribute to successive stages of the review.

```{r explain2}
reviewed <- reviewed |>
  explain(
    activity = "quality assurance",
    agent = "OpenRefine 3.10",
    used = "doi:10.2908/NAMA_10_GDP"
  )
```

The review history is therefore

``` text
circumference_candidate
            ↓
circumference_review_1
            ↓
circumference_review_2
```

## Finalising a review

Once the review is complete, the current reviewed values become the new
candidate values. The review history is retained, allowing future review rounds
to continue from the released version while preserving the complete review
trail.

```{r finalise}
reviewed <- reviewed |>
  finalise_review()
```

The review history is preserved, while future review rounds begin from the
finalised candidate values.

```{r names-finalised}
names(reviewed)
```

```{r finalised}
head(reviewed, 6)
```

The complete review algebra therefore consists of four operations: create
reviewable `claims`, allocate `review` rounds, document (`explain`) each review,
and promote the current review into the next candidate version.

## Complete workflow

```{r workflow }
claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
) |>
  review("circumference")

claims$circumference_review_1[1] <- 31

claims <- claims |>
  explain(
    activity = "manual review",
    agent = person("Jane", "Doe", role = "rev"),
    used = "doi:10.5281/zenodo.1234567"
  ) |>
  review("circumference")

claims$circumference_review_2[1] <- 32

claims <- claims |>
  explain(
    activity = "quality assurance",
    agent = "OpenRefine 3.10",
    used = "doi:10.2908/NAMA_10_GDP"
  ) |>
  finalise_review()
```

## Garamantes example

The review algebra is independent of the application domain. The same workflow
can be applied to statistical observations, archival metadata, cultural heritage
collections, or machine-generated annotations.

```{r garamantas}
garamantas <- data.frame(
  resources = c("https://garamantas.lv/en/file/475833",
                "https://garamantas.lv/en/file/471397",
                "https://garamantas.lv/en/file/475825"),
  instance_of = rep("photograph", 3),
  depicts  = rep("building", 3)
)

garamantas_claims <- claims_df(
    garamantas,
    scope_var = "instance_of",
    subject_var = "resources"
  ) |>
  review(c("depicts")) 

garamantas_claims$depicts_review_1[2] <- "group of people"


```

Although the domain differs, the review algebra remains unchanged:

``` text
claims
    ↓
review
    ↓
explain
    ↓
finalise_review
```

```{r showgaramatas}
garamantas_reviewed <- garamantas_claims |>
  explain(
    activity = "expert review",
    agent = person("Alice", "Curator", role = "rev"),
    used = "doi:example001"
  ) |>
  finalise_review()

print(garamantas_reviewed)

attributes(garamantas_reviewed)
```

Although the application domain differs, the review algebra remains\
unchanged. `review()` creates successive review versions, `explain()`\
records their provenance, and `finalise_review()` promotes the current\
reviewed values into the next candidate version.

The aim of the **review** package is to increase the trustworthiness,\
reviewability, and reusability of tabular data before they are released or\
exchanged with collaborators. Rather than replacing existing data\
engineering workflows, it adds a lightweight semantic review layer that\
preserves both the review history and its provenance.

Once a review has been completed, the resulting data can be published,\
archived, or exchanged using existing metadata standards. The companion\
[dataset package](https://dataset.dataobservatory.eu/) provides an R-native
representation for describing and\
serialising released datasets, allowing reviewed data to be exchanged in a\
reproducible and interoperable form.
