---
title: "The Review Algebra"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Review Algebra}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(review)
```

## Motivation

Many data workflows require human review before publication. Official
statistics, digital humanities, archives, machine learning, and data engineering
all rely on iterative assessment of candidate values before they become part of
a released dataset.

The goal of **review** is to make semantic review explicit and reproducible.
Rather than overwriting existing values, each review round creates a new version
of one or more reviewable claims while preserving the previous versions and
their provenance. This review history can be inspected, reproduced, and extended
with additional review rounds.

## Creating reviewable claims

The review package provides four verbs for reproducible semantic review:

- `claims_df()` creates reviewable claims.
- `review()` allocates a review round and labels the intended review task.
- `explain()` records how the review was carried out.
- `finalise_review()` accepts the reviewed values as the current candidate
  claims while preserving the complete review history.

```{r claims_df}
claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
)

head(claims, n = 6)
```

A `claims_df` separates variables into three structural roles and one or more
reviewable variables.

- `Identifier` uniquely identifies each claim.
- `Scope` defines the context in which the claim is made.
- `Subject` identifies the entity being described.
- `Reviewable` variables contain the values that may change during semantic
  review.

The structural variables remain stable throughout the review process, while
reviewable variables may acquire successive reviewed versions.

Readers familiar with the tidy data principles will recognise this separation.
Structural variables play a role analogous to identifiers, dimensions, and
attributes in statistical data production, while the reviewable variables
correspond to measured values that are iteratively reviewed and improved. The
structural variables provide the context for interpretation and allow claims to
be grouped, filtered, or compared, whereas the reviewable variables capture the
semantic content that evolves during the review process.

```{r names}
names(claims)
```

Notice that `circumference` has become `circumference_candidate`. Candidate
values are the current working version of each reviewable variable and form the
starting point for the first review round.

## First review round

A review round allocates a new review column for a reviewable variable and
records a label describing the intended review task. The label serves as an
instruction before the review is carried out and later identifies the completed
review activity.

```{r review1}
reviewed <- claims |>
  review(
    "circumference",
    label = "Remeasure the circumference of each tree."
  )

attr(reviewed, "review_label")
```

The review algebra is independent of the review interface. Reviewers may use
whatever environment best suits the task, provided that the reviewed values are
written back into the allocated review columns.

The package deliberately does not prescribe how the review is carried out. For
example, the review may be

- a `dplyr::mutate()` pipeline,
- values imported from a reviewed CSV file,
- an interactive R console session,
- a Shiny application,
- an Excel or LibreOffice spreadsheet,
- an OpenRefine workflow.

For illustration we edit one value directly.

```{r review1b}
reviewed$circumference_review_1[1] <- 31
```

Once the review values have been entered, the review round can be documented.

```{r explain1}
reviewed <- reviewed |>
  explain(
    activity = "manual review",
    agent = person("Jane", "Doe", role = "rev"),
    used = "doi:10.5281/zenodo.1234567",
    comment = "Verified against the laboratory notebook."
  )

attr(reviewed, "prov_comment")
```

`explain()` records how the review was performed, who performed it, which
evidence or resources informed the review, and optional reviewer comments
documenting how the review task was interpreted or why particular decisions were
made.

Provenance is stored separately from the reviewed values, allowing the same
review data to be interpreted or exported in different provenance models.

Readers familiar with FAIR data principles, reproducible research, or
statistical and digital heritage workflows may recognise a common problem: as
data are cleaned, harmonised, and reviewed, an increasing share of the knowledge
about *why* particular decisions were made remains only in the analyst's or
curator's head. The resulting datasets may be reusable, but the review process
itself is often difficult to inspect, reproduce, or audit.

The review algebra aims to record as much of this review history as possible
while requiring as little additional documentation effort as possible. By
separating reviewed values from the provenance of the review, it records which
**agents** (people, software, or AI models) performed which **activities**, and
what **entities** (datasets, publications, files, or other resources) informed
those decisions.

`review()` and `explain()` deliberately separate the description of a review
task from the documentation of its execution.

- A review **label** describes what reviewers are asked to do. It therefore
  serves as a task description before the review and as the label of the
  completed review activity afterwards.

- A reviewer **comment** explains how the task was interpreted or why particular
  review decisions were made.

This lightweight provenance layer follows the concepts of the PROV data model
without requiring users to work directly with RDF or ontologies. It can later be
serialised to PROV-O or aligned with standards such as SDMX, DataCite, or
archival provenance models. The objective is to improve the **reviewability**,
**auditability**, **reproducibility**, and ultimately the **trustworthiness**
and **reusability** of reviewed data while remaining compatible with ordinary R
data frames.

```{r attributes1}
attr(reviewed, "prov_activity")

attr(reviewed, "prov_agent")

attr(reviewed, "prov_used")
```

## Complete Review Workflow

```{r workflow }
claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
) |>
  review("circumference")

claims$circumference_review_1[1] <- 31

claims <- claims |>
  explain(
    activity = "manual review",
    agent = person("Jane", "Doe", role = "rev"),
    used = "doi:10.5281/zenodo.1234567"
  ) |>
  review("circumference")

claims$circumference_review_2[1] <- 32
claims <- claims |>
  explain(
    activity = "quality assurance",
    agent = "OpenRefine 3.10",
    used = "doi:10.2908/NAMA_10_GDP"
  ) |>
  finalise_review()
```

## Adding subsequent rounds of review

Each review round is allocated from the current candidate values. After
`finalise_review()`, those candidate values become the accepted result of the
previous review, allowing further review rounds to continue from the updated
claims.

Consider this second-round review:

```{r review2}
reviewed <- reviewed |>
  review("circumference")
```

Again, the review itself may take place using any suitable workflow.

```{r review2value}
reviewed$circumference_review_2[1] <- 32
```

Each review round has its own provenance. Different reviewers, software, or
evidence can therefore contribute to successive stages of the review.

```{r explain2}
reviewed <- reviewed |>
  explain(
    activity = "quality assurance",
    agent = "OpenRefine 3.10",
    used = "doi:10.2908/NAMA_10_GDP"
  )
```

The review history is therefore

``` text
circumference_candidate
            ↓
circumference_review_1
            ↓
circumference_review_2
```

## Finalising a review

Once the review is complete, the current reviewed values become the new
candidate values. The review history is retained, allowing future review rounds
to continue from the released version while preserving the complete review
trail.

```{r finalise}
reviewed <- reviewed |>
  finalise_review()
```

The review history is preserved, while future review rounds begin from the
finalised candidate values.

```{r names-finalised}
names(reviewed)
```

```{r finalised}
head(reviewed, 6)
```

The complete review algebra therefore consists of four operations: create
reviewable `claims`, allocate `review` rounds, document (`explain`) each review,
and promote the current review into the next candidate version.

## Dual structure and complete workflow

The review algebra can be viewed as a dual extension of tidy data.

An ordinary tidy data frame represents observations in rows and variables in
columns. The tidyverse provides an expressive algebra for transforming these
observations through operations such as `mutate()`, `filter()`, `select()`,
`join()`, and `pivot_*()`.

The review algebra introduces a second algebra that is structurally aligned with
the tidy data algebra. The upper layer contains successive semantic states of
the same claim vectors (reviewable variables), while the lower layer records
provenance describing with statements about the transitions between those states
how the transition took place.

```         
                 REVIEW MATRIX

┌──────────────────────────────────────────────────────┐
│ Upper layer: review states                           │
│                                                      │
│ candidate → review 1 → review 2 → ... → reviewed     │
└──────────────────────────────────────────────────────┘
┌──────────────────────────────────────────────────────┐
│ Lower layer: provenance                              │
│                                                      │
│ activity • agent • resources • comments              │
└──────────────────────────────────────────────────────┘
```

The two layers share the same structure. Every review column in the upper layer
has a corresponding provenance description in the lower layer. This shared
structure makes it possible to derive provenance automatically at different
levels of detail.

A single provenance statement may describe an entire review column:

> "The bibliographic information was verified by Librarian 1."

or, if required, the provenance can be refined to individual claims:

> "Rows 001, 004 and 005 were corrected after consulting manuscript B."

Because both layers share the same identifiers and ordering, column-level
provenance and cell-level provenance are different resolutions of the same
underlying model.

### Review sequencing

Review sequencing

A review column represents a semantic mutation of a reviewable variable.

Unlike ordinary `mutate()` operations, review mutations are ordered because
their sequence may affect the final result. For example, a two-stage imputation
or a sequence of authority-file reconciliations is generally not commutative.

For this reason, every review column belongs to a strictly ordered review chain.

A complete workflow may consist of several reviews or refinements.

```         
               review algebra

bibliography_candidate
          │
          ▼
bibliography_author_year
          │
          ▼
bibliography_publisher
          │
          ▼
bibliography_reviewed


              provenance algebra

author_year
    activity
    agent
    used
    ...

publisher
    activity
    agent
    used
```

The column names provide meaningful identifiers for the review stages, whereas
the package records their ordering internally. This allows review stages to be
named according to their intended purpose while preserving the temporal sequence
of the review.

Each review stage therefore has two complementary descriptions.

A `label` records the intended review activity before it is carried out. An
explanation records what actually happened during the review.

The label expresses the planned activity:

> "Verify authors and publication years."

whereas the explanation documents its execution in an optional comment:

> "Verified by Librarian 1 using the National Bibliography. Three publication
> years were corrected."

This distinction mirrors the distinction in provenance models between an
activity specification and the execution of that activity.

### Relation to Supply and Use and Input-Output Tables

Readers familiar with Supply and Use Tables (SUTs) or Symmetric Input-Output
Tables (SIOTs) may recognise a similar design pattern.

Supply and Use Tables (SUTs) and Symmetric Input-Output Tables (SIOTs) consist
of several matrices that describe different aspects of the same economy while
sharing a common ordering of industries and products. Because these matrices are
structurally aligned, aggregates such as gross domestic product (GDP) become
deterministic functions of the detailed accounts (f.e., total consumption, total
import.) Changes in one part of the system therefore propagate consistently
through the others.

The review algebra applies the same principle to semantic review. The upper
matrix records successive semantic states of a tidy data frame, while the lower
matrix records the provenance associated with those states. Because the two
matrices share the same structure, provenance can be generated at the level of
entire review stages or, when necessary, for individual reviewed claims.

Unlike many review systems, `review` does not store separate audit logs or
provenance graphs as primary data structures. Instead, it stores successive
semantic states of reviewable variables together with lightweight provenance
describing each review stage. Audit reports, provenance graphs, release notes,
and cell-level revision histories are deterministic projections of this
representation rather than independently maintained artefacts.

## Garamantas example

The review algebra is independent of the application domain. The same workflow
can be applied to statistical observations, archival metadata, cultural heritage
collections, or machine-generated annotations.

```{r garamantas}
garamantas <- data.frame(
  resources = c(
    "https://garamantas.lv/en/file/475833",
    "https://garamantas.lv/en/file/471397",
    "https://garamantas.lv/en/file/475825"
  ),
  instance_of = rep("photograph", 3),
  depicts = rep("building", 3)
)

garamantas_claims <- claims_df(
  garamantas,
  scope_var = "instance_of",
  subject_var = "resources"
) |>
  review(c("depicts"))

garamantas_claims$depicts_review_1[2] <- "group of people"
```

Although the domain differs, the review algebra remains unchanged:

``` text
claims_df()
      ↓
review()
      ↓
explain()
      ↓
finalise_review()
```

```{r showgaramatas}
garamantas_reviewed <- garamantas_claims |>
  explain(
    activity = "expert review",
    agent = person("Alice", "Curator", role = "rev"),
    used = "doi:example001"
  ) |>
  finalise_review()

print(garamantas_reviewed)

attributes(garamantas_reviewed)
```

## Conclusion

Although the application domain differs, the review algebra remains unchanged.
`review()` creates successive review versions, `explain()`records their
provenance, and `finalise_review()` promotes the current reviewed values into
the next candidate version.

The aim of the **review** package is to increase the trustworthiness,
reviewability, and reusability of tabular data before they are released or
exchanged with collaborators. Rather than replacing existing data engineering
workflows, it adds a lightweight semantic review layer that preserves both the
review history and its provenance. The package achieves this with a dual
extension of the `tidyverse` data model for iterative semantic review of tabular
data.

The package extends ordinary tidy workflows with review states and provenance
while preserving compatibility with familiar data-frame operations. It provides
a vectorised review workspace for creating, reviewing,documenting, and
finalising semantic claims without prescribing how reviews are carried out,
allowing review activities to be performed in R or external tools such as
spreadsheets, databases, and interactive review interfaces.

Although it is not an ontology toolkit, its design is informed by provenance,
policy, and statistical quality-control models, enabling review histories to be
audited, serialised, and exchanged in a reproducible form (You can read more
about this in the [From Review Algebra to Provenance
Modelling](https://review.dataobservatory.eu/articles/prov.html) vignette.)

Once a review has been completed, the resulting data can be published, archived,
or exchanged using existing metadata standards. The companion [dataset
package](https://dataset.dataobservatory.eu/) provides an R-native
representation for describing and serialising released datasets, allowing
reviewed data to be exchanged in a reproducible and interoperable form.
