---
title: "From Review Algebra to Provenance Modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{From Review Algebra to Provenance Modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(review)
```

### Review algebra is implementation-independent

The same review object can be edited by

-  dplyr

-  Excel

-  Shiny

-  OpenRefine

### Review rounds as provenance activities

Show

```         
candidate ↓ review_1 ↓ review_2
```

and explain that every review round corresponds to one provenance activity.

### Provenance attributes

For example

```         
attr(gdp, "prov_activity")  candidate  review_1  review_2 download   impute    forecast
```

```         
attr(gdp, "prov_agent")
```

```         
attr(gdp, "prov_used")
```

### Relationship to PROV

Without implementing RDF.

Simply explain

| Review algebra           | PROV           |
|--------------------------|----------------|
| candidate/review columns | Entity         |
| prov_activity            | Activity       |
| prov_agent               | Agent          |
| prov_used                | used           |
| review chain             | wasGeneratedBy |

The important point is

> 
>
> The package stores provenance in a lightweight R representation that can later be serialised into PROV or other provenance models.

### Outlook

Mention

-  RDF

-  PROV-O

-  RiC

-  SDMX

-  DataCite

without implementing them.
