---
title: "tidyverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyverse}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(review)
```

The `review` package offers a dual extension of the tidyverse data model for
iterative semantic review of tabular data. In the introductory vignette, [The
Review Algebra](https://review.dataobservatory.eu/articles/intro.html), we
introduce the four verbs that support semantic review. The second vignette,
[From Review Algebra to Provenance
Modelling](https://review.dataobservatory.eu/articles/prov.html), explained how
review rounds can be interpreted as lightweight provenance activities.

This vignette focuses on how the review algebra fits naturally into ordinary tidyverse workflows. The package extends ordinary tidy workflows with review states and provenance while preserving compatibility with familiar data-frame
operations. The tidyverse provides an elegant relational algebra for
transforming data frames. Following the principles of tidy data, observations,
dimensions and attributes are represented explicitly, while many analytical
assumptions, quality judgements and review decisions remain implicit in the
analyst's workflow. During interactive data analysis this is often exactly the
right design.

However, once data leave the analyst's workspace, those implicit decisions
become difficult for others to inspect or reproduce. This is particularly
important when review involves external collaborators, domain experts,
statistical production pipelines, or released datasets that should satisfy the
FAIR principles of reusability and reproducibility.

Several packages attempt to bridge this gap by serialising data together with
metadata. In practice, however, separating data from their metadata often
creates another maintenance problem: unless both are updated together, they
gradually drift apart.

The review package takes a different approach. Rather than introducing a new
data representation, it extends ordinary tidyverse data frames with a
lightweight review layer that remains attached to the data throughout the
workflow. Review rounds, reviewer annotations and provenance therefore travel
with the data, while the object continues to behave like a familiar tidy data
frame.

This design aims to bridge the gap between interactive data analysis and the
exchange of reviewable, reproducible and trustworthy data without requiring
users to abandon existing tidyverse workflows.

## Review as a pause in a tidy workflow

Tidyverse pipelines usually move directly from one transformation to the next.
In many real workflows, however, this is not enough. A value may need to be
checked by a human reviewer, reconciled in another tool, or returned from an
external process before the workflow can continue.

The `review` package adds this missing in-between state.

``` text
data
  ↓
claims_df()
  ↓
review()
  ↓
external review
  ↓
explain()
  ↓
finalise_review()
```

`review()` does not perform the review. It allocates a review column and allows
the workflow to pause.

```{r}
claims <- claims_df(
  Orange,
  scope_var = "age",
  subject_var = "Tree"
)

reviewed <- claims |>
  review(
    "circumference",
    label = "Check the circumference values."
  )

names(reviewed)
```

The reviewer may now edit the allocated review column. This can happen in many
ways: in R (either via an interactive data entry on the consol, or via a
mutation of the assigned review column), in a spreadsheet, in OpenRefine, in
Shiny, or in another external process.

## Review with mutate

A review can be expressed as an ordinary `mutate()` step when the review decision is reproducible in code.

```{r mutatereview}
library(dplyr)
reviewed <- reviewed |>
  dplyr::mutate(
    circumference_review_1 = 
      dplyr::if_else(
         claim_id == 1,
          31,
          circumference_review_1
    )
  )
```

The review round can then be documented.

```{r mutateexplain}
reviewed <- reviewed |>
  explain(
    activity = "coded correction",
    agent = "dplyr::mutate()",
    comment = "First circumference value corrected in code."
  )
```

## Review with an external file

A review may also leave R entirely. For example, the allocated review column can
be written to a temporary CSV file, edited elsewhere, and read back into the
same review object.

```{r extfilereview}
review_file <- tempfile(fileext = ".csv")

write.csv(
  reviewed[c("claim_id", "circumference_review_1")],
  review_file,
  row.names = FALSE
)
```

In a real workflow, the CSV file could be reviewed in a spreadsheet or another
curation tool. Here we simulate the edited file.

```{r extfilereview2}
external_review <- read.csv(review_file)
external_review$circumference_review_1[2] <- 34
write.csv(external_review, review_file, row.names = FALSE)

external_review <- read.csv(review_file)
```

The reviewed values are then joined back by the claim identifier.

```{r relocate}
reviewed <- reviewed |>
  select(-circumference_review_1) |>
  left_join(external_review, by = "claim_id") |>
  relocate(
    circumference_review_1,
    .after = circumference_candidate
  )
```

The same review round can still be explained.

```{r spreadsheet}
reviewed <- reviewed |>
  explain(
    activity = "spreadsheet review",
    agent = person("Jane", "Doe", role = "rev"),
    used = review_file,
    comment = "Reviewed values were read back from a CSV file."
  )
```

## Review in the R console

For small examples, review can also be interactive. A reviewer can inspect the object and overwrite one or more values directly.

```{r}
reviewed$circumference_review_1[3] <- 37
```

This is not less valid than a spreadsheet or a pipeline. What matters is that the reviewed value is stored in the allocated review column and the review is documented with `explain()`.

## Scope of the review

The review algebra works best on single vectors. In this case, the review is a
row-by-row revision of the validity of some claim on a variable. Technically, it
would be possible to review several variables at one time, which would mean the
revision of interactions.

```{r varscope}
data.frame(
  rowid = c(1, 2),
  title = c("Semantics", "Revisions"),
  author = c("Jane Doe", "Alice Cooper"),
  year = c("2026", "2021")
) %>%
  claims_df(scope_var = "rowid", subject_var = "title")
```

```{r}
library(dplyr)

data.frame(
  rowid = c(1, 2),
  title = c("Semantics", "Revisions"),
  author_year = c("Jane Doe 2026", "Alice Cooper 2021")
) |>
  claims_df(scope_var = "rowid", subject_var = "title") |>
  review(
    review_var = "author_year",
    label = "Check the author and year of the articles"
  ) |>
  dplyr::mutate(author_year_review_1 = ifelse(
    test = author_year_review_1 == "Alice Cooper 2021",
    yes = "Alice Cooper 2022", 
    no = author_year_review_1)
    ) |>
  explain(
    activity = "bibliography check",
    agent = person("Librarian"),
    comment = "Alice Cooper seems to published only in 2022"
  )
```

## Finalising the review

Once the reviewed values are accepted, `finalise_review()` promotes the current
review values back into the candidate columns.

```{r}
accepted <- reviewed |>
  finalise_review()

accepted$circumference_candidate[1:3]
```

The review history is preserved.

```{r}
names(accepted)
attr(accepted, "prov_activity")
attr(accepted, "prov_agent")
attr(accepted, "prov_comment")
```

## A tidy annotation layer

The review algebra is designed to remain close to ordinary data-frame workflows.
It does not replace `dplyr`; it adds review state, provenance, and review
history around transformations that can still be inspected with ordinary R
tools.

This makes `review` similar in spirit to `dataset_df`: it adds a semantic layer
while remaining compatible with the tidyverse data-frame model. The larger
design idea is not to replace tidy data, but to make the assumptions and review
decisions around tidy data more explicit.

This works best for vector-wise transformations such as `mutate()`,
`transmute()`, `filter()`, `select()`, `relocate()`, `fill()`.

Operations that collapse or reshape the data, such as `summarise()` or some
`pivot_*()` workflows, may change the identity of claims and therefore need more
explicit handling.

## Conclusion

The `tidyverse` already provides an algebra that is frequently used to construct
inference workflows. Data are filtered, reconciled, joined, transformed,
aggregated, and enriched to derive new semantic statements. These inference
steps are often deterministic, but their meaning and justification remain
largely implicit in the analysis code and the analyst's expertise.

The \`review\` package introduces a dual extension of the tidy algebra that
makes one class of these semantic transformations—human review—explicit. Rather
than merely producing a new data frame, it records the semantic states of
reviewable variables together with the provenance of the transformations that
produced them.

Human review is only the first application of this dual algebra. The same
computational model naturally extends to semantic inference. Rule-based
reasoning, statistical imputation, authority reconciliation, ontology alignment,
and AI-assisted annotation all produce successive semantic states that can be
represented using the same review algebra and provenance layer. In this broader
view, review is simply one specialised inference operator, distinguished by the
fact that the transformation is performed or validated by a human agent.
